---
title: "00 Download GDC counts and preprocessing"
author: ''
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
# Project root: works when (1) knitting or (2) running code in R (e.g. copy-paste)
# Run this chunk first if you run code interactively.
cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
if (!is.null(path) && length(path) > 0) {
  project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
} else {
  wd <- getwd()
  project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = project_root)
  knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 60))
}
# Wrap R output in PDF so long lines don't overflow
options(width = 70)
```

# Setup

Download TCGA GDC RNA-seq count files using a cart manifest, then preprocess into a merged count matrix.  
Data: [GDC Portal](https://portal.gdc.cancer.gov/).

**Running in R (copy-paste):** Set working directory to project root first, e.g. `setwd("/path/to/Tutorial")`, then run the **paths** chunk (and **packages**), then the rest in order. Or run from `Script/` — paths will still resolve.

```{r paths}
# Absolute paths from project root (same logic as setup — run setup first, or this defines project_root)
if (!exists("project_root")) {
  cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
  path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
  if (!is.null(path) && length(path) > 0) {
    project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
  } else {
    wd <- getwd()
    project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
  }
}
DATA_DIR    <- file.path(project_root, "Data")
COUNTS_DIR  <- file.path(DATA_DIR, "gdc_counts")
MANIFEST    <- file.path(DATA_DIR, "gdc_manifest.2026-03-12.122540.txt")
GDC_DATA_URL <- "https://api.gdc.cancer.gov/data"

COUNT_MATRIX_FILE <- file.path(paste0(DATA_DIR,"/Input"), "count_matrix.rds")
GENE_INFO_FILE    <- file.path(paste0(DATA_DIR,"/Input"), "gene_info.rds")
SAMPLES_FILE      <- file.path(paste0(DATA_DIR,"/Input"), "samples.rds")
```

# Packages

```{r packages}
library(httr)
library(readr)
library(dplyr)
library(tibble)
```

# 1. Download GDC counts

Create the output directory and read the manifest.

```{r download-setup}
dir.create(COUNTS_DIR, showWarnings = FALSE, recursive = TRUE)
manifest <- read_tsv(MANIFEST, show_col_types = FALSE)
n_files  <- nrow(manifest)
message("Manifest: ", n_files, " files → ", COUNTS_DIR)
```

Download each file by GDC file UUID (skips existing files with matching MD5).

```{r download-function}
download_one <- function(file_id, filename, expected_md5 = NULL) {
  url  <- paste(GDC_DATA_URL, file_id, sep = "/")
  dest <- file.path(COUNTS_DIR, filename)
  check_md5 <- !is.null(expected_md5) && !is.na(expected_md5)

  if (file.exists(dest)) {
    if (check_md5) {
      actual <- as.character(tools::md5sum(dest))
      if (!is.na(actual) && identical(actual, as.character(expected_md5))) {
        message("[skip] ", filename, " (already exists, MD5 OK)")
        return(invisible(TRUE))
      }
    } else {
      message("[skip] ", filename, " (already exists)")
      return(invisible(TRUE))
    }
  }

  # Omit progress() when knitting to avoid stray "|" and one-line-per-page in PDF
  prog <- if (interactive() && !knitr::is_latex_output()) progress() else list()
  resp <- GET(url, write_disk(dest, overwrite = TRUE), prog)
  if (http_error(resp)) {
    warning("Failed: ", filename, " - ", status_code(resp))
    if (file.exists(dest)) try(file.remove(dest))
    return(invisible(FALSE))
  }

  if (check_md5) {
    actual <- as.character(tools::md5sum(dest))
    if (is.na(actual) || actual != as.character(expected_md5)) {
      warning("MD5 mismatch or unreadable for ", filename)
      return(invisible(FALSE))
    }
  }
  message("[ok] ", filename)
  invisible(TRUE)
}
```

```{r download-run}
for (i in seq_len(n_files)) {
  with(manifest[i, ], download_one(id, filename, md5))
}
message("Download done → ", normalizePath(COUNTS_DIR))
```

# 2. Preprocessing

Find count files and build a single count matrix (genes × samples).

```{r find-files}
pattern <- "\\.rna_seq\\.augmented_star_gene_counts\\.tsv$"
count_files <- list.files(COUNTS_DIR, pattern = pattern, full.names = TRUE)
if (length(count_files) == 0) stop("No count files found in ", COUNTS_DIR)
length(count_files)
```

```{r read-helper}
read_one_counts <- function(path, count_col = "unstranded") {
  d <- read_tsv(path, comment = "#", show_col_types = FALSE)
  if (!count_col %in% names(d)) stop("Column not found: ", count_col)
  d <- d[, c("gene_id", "gene_name", "gene_type", count_col)]
  d[[count_col]] <- as.numeric(d[[count_col]])
  d
}

gene_template <- read_one_counts(count_files[1])
gene_info <- gene_template %>%
  select(gene_id, gene_name, gene_type) %>%
  filter(!startsWith(gene_id, "N_"))
```

```{r build-matrix}
sample_ids <- character(length(count_files))
count_list <- vector("list", length(count_files))

for (i in seq_along(count_files)) {
  f <- count_files[i]
  sample_id <- sub("\\.rna_seq\\.augmented_star_gene_counts\\.tsv$", "", basename(f))
  sample_ids[i] <- sample_id
  d <- read_one_counts(f) %>% filter(!startsWith(gene_id, "N_"))
  count_list[[i]] <- setNames(d[["unstranded"]], d[["gene_id"]])
}

genes <- gene_info$gene_id
count_matrix <- do.call(cbind, lapply(count_list, function(x) x[genes]))
count_matrix[is.na(count_matrix)] <- 0
rownames(count_matrix) <- genes
colnames(count_matrix) <- sample_ids
```

```{r samples}
samples <- tibble(
  sample_id = sample_ids,
  file = basename(count_files)
)
head(samples)
```

```{r save}
dir.create(paste0(DATA_DIR,"/Input"), showWarnings = FALSE, recursive = TRUE)
saveRDS(count_matrix, COUNT_MATRIX_FILE)
saveRDS(gene_info,   GENE_INFO_FILE)
saveRDS(samples,     SAMPLES_FILE)
```

# Summary

**Done.** Count matrix: `r nrow(count_matrix)` genes × `r ncol(count_matrix)` samples.

Outputs in `Data/`:

- `count_matrix.rds` — raw counts (genes × samples)
- `gene_info.rds` — gene_id, gene_name, gene_type
- `samples.rds` — sample_id, file
