---
title: "01 Preprocessing"
author: ''
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
# Project root (same pattern as 00_download_gdc_counts.Rmd)
cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
if (!is.null(path) && length(path) > 0) {
  project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
} else {
  wd <- getwd()
  project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = project_root)
  knitr::opts_chunk$set(
    echo = TRUE,
    message = TRUE,
    warning = TRUE,
    tidy = TRUE,
    tidy.opts = list(width.cutoff = 60)
  )
}
# Wrap R output in PDF so long lines don't overflow
options(width = 70)
```

# Setup

Read GDC cart metadata JSON, clean into a dataframe, and save to `Data/Input/`.

```{r paths}
if (!exists("project_root")) {
  cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
  path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
  if (!is.null(path) && length(path) > 0) {
    project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
  } else {
    wd <- getwd()
    project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
  }
}
DATA_DIR      <- file.path(project_root, "Data")
INPUT_DIR     <- file.path(DATA_DIR, "Input")
RESULTS_DIR   <- file.path(project_root, "Results", "01_Preprocessing")
METADATA_JSON <- file.path(DATA_DIR, "metadata.cart.2026-03-12.json")
CLINICAL_TSV  <- file.path(DATA_DIR, "clinical.cart.2026-03-12", "clinical.tsv")
SAMPLE_TSV    <- file.path(DATA_DIR, "biospecimen.cart.2026-03-12", "sample.tsv")
METADATA_RDS    <- file.path(RESULTS_DIR, "metadata.rds")
CLINICAL_RDS    <- file.path(RESULTS_DIR, "clinical.rds")
SAMPLE_RDS      <- file.path(RESULTS_DIR, "sample.rds")
COUNT_MATRIX_RDS <- file.path(INPUT_DIR, "count_matrix.rds")
```

# Packages

```{r packages}
library(jsonlite)
library(readr)
library(dplyr)
library(tibble)
```

# Helper for missing JSON fields

```{r helper}
`%||%` <- function(x, y) if (is.null(x)) y else x
```

# Read metadata JSON

```{r read-json}
raw <- read_json(METADATA_JSON, simplifyVector = FALSE)
length(raw)
```

# Clean into a dataframe

Flatten top-level file fields and the first `associated_entity` (case/aliquot) per file. Nested `analysis` and `annotations` are omitted for a simple table.

```{r clean-df}
rows <- lapply(raw, function(x) {
  ae <- if (length(x$associated_entities) > 0) x$associated_entities[[1]] else list()
  data.frame(
    file_id               = x$file_id %||% NA_character_,
    file_name             = x$file_name %||% NA_character_,
    submitter_id          = x$submitter_id %||% NA_character_,
    data_format           = x$data_format %||% NA_character_,
    data_category         = x$data_category %||% NA_character_,
    data_type             = x$data_type %||% NA_character_,
    file_size             = as.integer(x$file_size %||% NA_integer_),
    md5sum                = x$md5sum %||% NA_character_,
    state                 = x$state %||% NA_character_,
    platform              = x$platform %||% NA_character_,
    experimental_strategy = x$experimental_strategy %||% NA_character_,
    access                = x$access %||% NA_character_,
    case_id               = ae$case_id %||% NA_character_,
    entity_submitter_id   = ae$entity_submitter_id %||% NA_character_,
    entity_type           = ae$entity_type %||% NA_character_,
    entity_id             = ae$entity_id %||% NA_character_,
    workflow_type         = (if (!is.null(x$analysis)) x$analysis$workflow_type
                            else NULL) %||% NA_character_,
    stringsAsFactors      = FALSE
  )
})

metadata <- bind_rows(rows) %>%
  as_tibble() %>%
  mutate(sample_submitter_id = substr(entity_submitter_id, 1, 16))

dim(metadata)
head(metadata)
```

# Read clinical TSV, keep selected columns, rename, and remove duplicates

```{r read-clinical, tidy = FALSE}
clinical_raw <- read.delim(CLINICAL_TSV)
clinical <- clinical_raw %>%
  select(
    project.project_id,
    cases.case_id,
    cases.submitter_id,
    demographic.age_at_index,
    demographic.gender,
    demographic.race,
    demographic.vital_status,
    diagnoses.ajcc_pathologic_stage,
    diagnoses.ajcc_pathologic_t
  ) %>%
  rename(
    project_id           = project.project_id,
    case_id              = cases.case_id,
    submitter_id         = cases.submitter_id,
    age         = demographic.age_at_index,
    gender               = demographic.gender,
    race                 = demographic.race,
    vital_status         = demographic.vital_status,
    ajcc_pathologic_stage = diagnoses.ajcc_pathologic_stage,
    ajcc_pathologic_t    = diagnoses.ajcc_pathologic_t
  ) %>%
  mutate(has_missing = if_any(everything(), ~ . == "'--")) %>%
  group_by(case_id) %>%
  filter(
    n() == 1L |                                                      # case_id occurs once: keep it
    (n() > 1L & !has_missing) |                                     # case_id occurs multiple times: keep rows without '--'
    (n() > 1L & sum(!has_missing) == 0L & row_number() == 1L)        # case_id occurs multiple times and all contain '--: keep at least one row
  ) %>%
  ungroup() %>%
  select(-has_missing) %>%
  distinct()
message("Clinical: ", nrow(clinical_raw), " rows → ", nrow(clinical), " after selecting columns, cleaning duplicated case_id (drop '-- when possible, keep >= 1 per case_id), and distinct")
```

# Read sample TSV

```{r read-sample}
sample <- read.delim(SAMPLE_TSV) %>%
  dplyr::select(project.project_id, cases.case_id, samples.tissue_type, samples.submitter_id) %>%
  rename(
    project_id  = `project.project_id`,
    case_id     = `cases.case_id`,
    tissue_type = `samples.tissue_type`,
    sample_submitter_id = `samples.submitter_id`
  ) %>%
  filter(sample_submitter_id %in% metadata$sample_submitter_id)
  
dir.create(RESULTS_DIR, showWarnings = FALSE, recursive = TRUE)
saveRDS(sample, SAMPLE_RDS)

# Add clinical into sample, then map to entity_submitter_id via sample_submitter_id
clinical_one_per_case <- clinical %>% distinct(case_id, project_id, .keep_all = TRUE)
clinical <- sample %>%
  distinct(project_id, case_id, sample_submitter_id, .keep_all = TRUE) %>%
  left_join(clinical_one_per_case, by = c("case_id", "project_id"))

clinical <- metadata %>%
  select(entity_submitter_id, sample_submitter_id, case_id) %>%
  left_join(clinical, by = c("sample_submitter_id", "case_id")) %>%
  distinct(entity_submitter_id, .keep_all = TRUE)
```

# Add tissue_type from sample into metadata

```{r add-tissue-type}
sample <- sample %>% select(sample_submitter_id, tissue_type) %>% distinct(sample_submitter_id, .keep_all = TRUE)
metadata <- metadata %>% left_join(sample, by = "sample_submitter_id")
```

# Read count matrix and rename columns to entity_submitter_id

```{r read-count-matrix}
count_matrix <- readRDS(COUNT_MATRIX_RDS)
metadata$file_id_clean = gsub(".rna_seq.augmented_star_gene_counts.tsv", "",metadata$file_name)
idx <- match(colnames(count_matrix), metadata$file_id_clean)
colnames(count_matrix) <- ifelse(is.na(idx), colnames(count_matrix), metadata$entity_submitter_id[idx])
count_matrix[1:5, 1:5]
```

# Save to Results/01_Preprocessing

```{r save}
saveRDS(metadata,     METADATA_RDS)
saveRDS(clinical,     CLINICAL_RDS)
saveRDS(count_matrix, file.path(RESULTS_DIR, "count_matrix.rds"))
```

**Done.**

- Metadata: `r nrow(metadata)` rows x `r ncol(metadata)` columns -> `Results/01_Preprocessing/metadata.rds`
- Clinical: `r nrow(clinical)` rows -> `Results/01_Preprocessing/clinical.rds`
- Count matrix: `r nrow(count_matrix)` genes x `r ncol(count_matrix)` samples (cols = entity_submitter_id) -> `Results/01_Preprocessing/count_matrix.rds`
