---
title: "02 Differential expression (limma)"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
if (!is.null(path) && length(path) > 0) {
  project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
} else {
  wd <- getwd()
  project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = project_root)
  knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
}
options(width = 70)
```

# Setup

Breast cancer (Tumor) tissues were compared to Normal tissues. Differential expression was performed with **limma** (voom); genes with **|logFC| > 1.0** and **adj.P.Val < 0.05** were considered differentially expressed.

```{r paths}
if (!exists("project_root")) {
  cur <- tryCatch(knitr::current_input(), error = function(e) NULL)
  path <- if (is.null(cur)) NULL else if (is.character(cur)) cur else cur[["path"]]
  if (!is.null(path) && length(path) > 0) {
    project_root <- normalizePath(file.path(dirname(path[1]), ".."), mustWork = FALSE)
  } else {
    wd <- getwd()
    project_root <- if (dir.exists(file.path(wd, "Data"))) wd else normalizePath(file.path(wd, ".."), mustWork = FALSE)
  }
}
INPUT_DIR    <- file.path(project_root, "Results", "01_Preprocessing")
RESULTS_DIR  <- file.path(project_root, "Results", "02_DEG")
COUNT_MATRIX_RDS <- file.path(INPUT_DIR, "count_matrix.rds")
GENE_INFO_RDS     <- file.path("Data/Input", "gene_info.rds")
CLINICAL_RDS    <- file.path(INPUT_DIR, "clinical.rds")
```

# Packages

```{r packages}
library(dplyr)
library(tibble)
library(edgeR)
library(limma)
```

# Read data

```{r read-data}
count_matrix <- readRDS(COUNT_MATRIX_RDS)
clinical     <- readRDS(CLINICAL_RDS)
dim(count_matrix)
dim(clinical)
```

# Tumor vs Normal: design

```{r de-design}
# Matrix design https://www.bioconductor.org/packages//release/workflows/vignettes/RNAseq123/inst/doc/designmatrices.html
count_de <- count_matrix[, clinical$entity_submitter_id, drop = FALSE]
clinical$tissue_type <- factor(clinical$tissue_type, levels = c("Normal", "Tumor"))
design <- model.matrix(~ 0 + tissue_type, data = clinical)
colnames(design) <- gsub("tissue_type", "", colnames(design))
```

# Filter lowly expressed genes (edgeR)

As in the [limma/edgeR workflow](https://stemangiola.github.io/zhejiang2020/articles/limmaWorkflow.html), genes without worthwhile counts are removed before normalisation and voom to improve the mean-variance estimate and reduce unnecessary tests.

```{r de-filter-genes}
group <- clinical$tissue_type
dge <- DGEList(counts = count_de, group = group)

gene_info = readRDS(GENE_INFO_RDS)
dge$genes = gene_info[match(rownames(dge), gene_info$gene_id), c("gene_id","gene_name", "gene_type")]

# Keep only protein_coding genes
dge <- dge[which(dge$genes$gene_type == "protein_coding"), , keep.lib.sizes = FALSE]
message("Protein-coding genes: ", nrow(dge), " / ", nrow(count_de))

# Transformations from the raw-scale
cpm <- cpm(dge)
lcpm <- cpm(dge, log=TRUE)
L <- mean(dge$samples$lib.size) * 1e-6
M <- median(dge$samples$lib.size) * 1e-6

# Removing genes that are lowly expressed
keep <- filterByExpr(dge, group = group)
dge <- dge[keep, , keep.lib.sizes = FALSE]
message("Genes retained after filterByExpr: ", nrow(dge), " / ", nrow(count_de))

lcpm.cutoff <- log2(10/M + 2/L)
nsamples <- ncol(dge)
samplenames <- substring(colnames(dge), 12, nchar(colnames(dge)))
par(mfrow=c(1,2))
plot(density(lcpm[,1]), lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, lwd=2)
}
lcpm <- cpm(dge, log=TRUE)
plot(density(lcpm[,1]), lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, lwd=2)
}
```

# TMM normalisation and limma voom

```{r de-voom}
dge <- calcNormFactors(dge, method = "TMM")

# the expression distribution of samples for normalised data
lcpm <- cpm(dge, log=TRUE)
boxplot(lcpm, las=2, main="")
title(main="Normalised data",ylab="Log-cpm")

v <- voom(dge, design, plot = TRUE)

fit <- lmFit(v, design)
```

# Contrasts, eBayes, and topTable

```{r de-decide}
contrasts <- makeContrasts(Tumor - Normal, levels = colnames(design))
vfit <- contrasts.fit(fit, contrasts = contrasts)
efit <- eBayes(vfit)
plotSA(efit, main = "Final model: Mean-variance trend")

Tumor_vs_normal <- topTable(efit, n = Inf)
# DEGs: filter by |logFC| > 1 and adj.P.Val < 0.05
de_genes <- Tumor_vs_normal[abs(Tumor_vs_normal$logFC) > 1 & Tumor_vs_normal$adj.P.Val < 0.05, ]
message("DEGs (|logFC| > 1, adj.P.Val < 0.05): ", nrow(de_genes))
```

# Volcano plot

```{r volcano}
res <- as.data.frame(Tumor_vs_normal)
res$sig <- "NotSig"
res$sig[res$logFC > 1 & res$adj.P.Val < 0.05] <- "Up"
res$sig[res$logFC < -1 & res$adj.P.Val < 0.05] <- "Down"
res$sig <- factor(res$sig, levels = c("Down", "NotSig", "Up"))

with(res, plot(logFC, -log10(adj.P.Val), pch = 20, main = "Volcano: Tumor vs Normal",
  xlab = "log2 fold change", ylab = "-log10(adj.P.Val)"))
abline(h = -log10(0.05), lty = 2)
abline(v = c(-1, 1), lty = 2)
with(subset(res, sig == "Up"), points(logFC, -log10(adj.P.Val), pch = 20, col = "red"))
with(subset(res, sig == "Down"), points(logFC, -log10(adj.P.Val), pch = 20, col = "blue"))
legend("topright", legend = c("Down", "Not sig", "Up"), pch = 20, col = c("blue", "black", "red"), bty = "n")
```

# Save

```{r save}
dir.create(RESULTS_DIR, showWarnings = FALSE, recursive = TRUE)
saveRDS(efit, file.path(RESULTS_DIR, "efit_limma.rds"))
saveRDS(vfit, file.path(RESULTS_DIR, "vfit_limma.rds"))
write.csv(Tumor_vs_normal, file.path(RESULTS_DIR, "Tumor_vs_normal.csv"), row.names = TRUE)
saveRDS(de_genes, file.path(RESULTS_DIR, "de_genes_limma.rds"))
message("Saved to ", RESULTS_DIR)
```
